---
title: "Timeseries modeling of Dynamic World probabilities"
output: 
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
date: "`r Sys.Date()`"
---

# References

Built on the [`{compositions}`](http://www.stat.boogaart.de/compositions/) package which is described by the authors in this [paper](https://sci-hub.se/10.1016/j.cageo.2006.11.017).

A handy primer on how to use this for linear modeling is [UsingCompositions.pdf](http://www.stat.boogaart.de/compositions/UsingCompositions.pdf). See, specifically, pg. 13 which demonstrates how to build a model where the response variable is compositional data, and predictor variable/s is/are real data (which is the case for us).

# Load libraries

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(geojsonsf)
library(sf)
library(compositions)
# library(robCompositions)
```

# Load & prepare DW data
```{r}
ptsWithProbs <- geojson_sf("data/labPtsWithDW.geojson")
```

## Go from 9 to 3 probabilities
Also, create convert time stamp to date.
```{r}
ptsWithProbsCollapsed <- ptsWithProbs %>% 
  mutate(nat = trees + snow_and_ice + crops +
           grass + shrub_and_scrub + bare,
         nonnat = built,
         wet = flooded_vegetation + water,
         date = as.Date(`system:time_start`/(1e3*60*60*24)), # day
         timeMillis = `system:time_start`,
         .keep = "unused")
```

# For a single point

```{r}
point <- ptsWithProbsCollapsed %>% filter(ptId == "77.4940_12.9205")
```

## Linear fit 

### Year 2020

Probabilities point cloud

```{r}
year = '2020'
point20 <- point %>% 
  filter(between(
    date, 
    as.Date(paste0(year, '-01-01')), 
    as.Date(paste0(year, '-12-31')))) %>% 
  drop_na(any_of(c("wet", "nat", "nonnat")))

pointX20 <- point20 %>% select(date) %>% 
  st_drop_geometry()
pointY20 <- point20 %>% select(nat, nonnat, wet) %>% 
  st_drop_geometry() %>% 
  acomp()
plot(pointY20)
```

Linear fit Y ~ X
```{r}
mylm20 <- lm(ilr(pointY20) ~ X, data = data.frame(X = pointX20$date))
summary(mylm20)
```

Model predictions, in composition space
```{r}
plot(ilrInv(predict(mylm20)), pch=20)
```

Prediction residuals, in composition space
```{r}
plot(ilrInv(resid(mylm20)))

# ellipses(mean(pointY), var(pointY))
# yPred <- ilrInv(predict(mylm, newdata=data.frame(X=pointX$date)))
# plot(yPred)

```

### Year 2022

Probabilities point cloud

```{r}
year = '2022'
point22 <- point %>% 
  filter(between(
    date, 
    as.Date(paste0(year, '-01-01')), 
    as.Date(paste0(year, '-12-31')))) %>% 
  drop_na(any_of(c("wet", "nat", "nonnat")))

pointX22 <- point22 %>% select(date) %>% 
  st_drop_geometry()
pointY22 <- point22 %>% select(nat, nonnat, wet) %>% 
  st_drop_geometry() %>% 
  acomp()
plot(pointY22)
```

Linear fit Y ~ X

```{r}
mylm22 <- lm(ilr(pointY22) ~ X, data = data.frame(X = pointX22$date))
summary(mylm22)
```

Model predictions
```{r}
plot(ilrInv(predict(mylm22)), pch=20)
```

Prediction residuals
```{r}
plot(ilrInv(resid(mylm22)))

# ellipses(mean(pointY), var(pointY))
# yPred <- ilrInv(predict(mylm, newdata=data.frame(X=pointX$date)))
# plot(yPred)
```

## Harmonic fit

### Year 2022
Harmonic fit Y ~ X + cos(2piX) + sin(2piX)
```{r}
xc <- cos(2*pi*point22$timeMillis)
xs <- sin(2*pi*point22$timeMillis)
mylmHarm22 <- lm(
  ilr(pointY22) ~ X + XC + XS, 
  data = data.frame(X = pointX22$date, XC = xc, XS = xs))
summary(mylmHarm22)
```

Model predictions
```{r}
plot(ilrInv(predict(mylmHarm22)), pch=20)
```

Prediction residuals
```{r}
plot(ilrInv(resid(mylmHarm22)))
```
