---
title: "Modeling Dynamic World probabilities as timeseries of compositional data vectors"
output: 
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
date: "`r Sys.Date()`"
---

# Overview

## Context
A pixel in a Dynamic World (DW) image captures probabilities 9 land cover categories. These probabilities at every pixel change as its land cover cycles through seasons each year. Modeling patterns in a timeseries of these probabilities can prove powerful in inferring more complex land cover types than are available in DW.

Using probability values of a single cover type (say, `grass`) for such timeseries modeling (to infer, say, different types of grasslands) is one way to approach this. But doing so ignores the neat mathematical structure offered by the fact that it is one component of a vector of probabilities, which change together while also summing to 1 at every time point, i.e., the probability vectors have the structure of compositional data vectors, and DW offers a timeseries of those at each pixel.

In our land cover mapping exercise identifying different types of open-canopy natural ecosystems in semi-arid India (roughly 2.5M sq. km.), **we wish to explore the compositional approach to modeling vegetation phenology using multiple linear regression on the DW timeseries** (along the lines of Nick Clinton's tutorial on harmonic modeling in the [Time Series Analysis in Earth Engine](https://www.youtube.com/watch?v=xQGNAYL7z80) tutorial). This would involve the following steps, which is achievable using the [`{compositions}`](http://www.stat.boogaart.de/compositions/) package in R:

* apply isometric logratio (ilr) transformation to each probability vector in a timeseries
* model landcover phenology by performing harmonic analysis, using multiple linear regression, on the ilr-transformed vectors.

We have attempted this below for one pixel at a time. To be able to visualize results and build our understanding, we have reduced the numerical complexity down to a 3 probabilities case, from the 9 probabilities offered by DW.

## How Earth Engine could help
We would like to perform this harmonic modeling on the DW dataset treated as a timeseries of compositional vector data of 9-long probability vectors. Using the results of this harmonic modeling (the coefficients, goodness of fit), we would like to build a classifier using one of the classifiers already available on Earth Engine to produce a map.

Specifically, **can there be a reducer which can perform multiple linear regression on the (ilr-transformed) compositional vectors as response variables**, returning model coefficients and goodness of fit measures?

# Additional resources

* An overview of the math behind linear regression of compositional data, isometric logratio: Wang et al. (2013), "Multiple linear regression modeling for compositional data", Neurocomputing, Volume 122, 25 December 2013, Pages 490-500. [(SciHub link)](https://sci-hub.se/10.1016/j.neucom.2013.05.025).
* Built on the [`{compositions}`](http://www.stat.boogaart.de/compositions/) package which is described by the authors in this [paper](https://sci-hub.se/10.1016/j.cageo.2006.11.017).
  * A handy primer on how to use this for linear modeling is [UsingCompositions.pdf](http://www.stat.boogaart.de/compositions/UsingCompositions.pdf). See, specifically, pg. 13 which demonstrates how to build a model where the response variable is compositional data, and predictor variable/s is/are real data (which is the case for us).

# Code to explore DW harmonic mideling for a single point

## Load libraries

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(geojsonsf)
library(sf)
library(compositions)
# library(robCompositions)
```

## Load & prepare DW data
```{r}
ptsWithProbs <- geojson_sf("data/labPtsWithDW.geojson")
head(ptsWithProbs)
```

### Go from 9 to 3 probabilities
Also, create convert time stamp to date.
```{r}
ptsWithProbsCollapsed <- ptsWithProbs %>% 
  mutate(nat = trees + snow_and_ice + crops +
           grass + shrub_and_scrub + bare,
         nonnat = built,
         wet = flooded_vegetation + water,
         date = as.Date(`system:time_start`/(1e3*60*60*24)), # day
         timeMillis = `system:time_start`,
         .keep = "unused")
```

## For a single point
Choose a single point to run a compositional linear model over. At-scale, this will have to happen for all points in the table (for land cover classifier training) and for every pixel in the image stack (for classifier prediction).
```{r}
chosenPoint <- ptsWithProbsCollapsed %>% filter(ptId == "76.3114_14.7652")
# 77.4940_12.9205: built
# 76.3114_14.7652: savanna_open
# 75.5558_12.2999: agri_hiBiomass
# 76.3071_14.7580: savanna_shrub
# 75.5625_12.2379: forest
chosenPoint
```

### Linear fit 
Since most (natural) land covers have a cycle of seasons every year, the phenology traced by the probability vector is assumed to have a period of 1 year.

#### Year: 2020
Choose a single year. Plot the probabilities point cloud in ternary space.

```{r}
year = '2020'
point20 <- chosenPoint %>% 
  filter(between(
    date, 
    as.Date(paste0(year, '-01-01')), 
    as.Date(paste0(year, '-12-31')))) %>% 
  drop_na(any_of(c("wet", "nat", "nonnat")))

point20X <- point20 %>% select(date) %>% 
  st_drop_geometry() # to save memory
point20Y <- point20 %>% select(nat, nonnat, wet) %>% 
  st_drop_geometry() %>% 
  acomp() # force the probability triplet to sum to 1 making it compositional data
plot(point20Y)
```

Simple linear fit Y ~ X
```{r}
simpleLm20 <- lm(ilr(point20Y) ~ X, data = data.frame(X = point20X$date))
summary(simpleLm20)
```

Model predictions, in composition space
```{r}
plot(ilrInv(predict(simpleLm20)), pch=20)
```

Prediction residuals, in composition space
```{r}
plot(ilrInv(resid(simpleLm20)))

# ellipses(mean(pointY), var(pointY))
# yPred <- ilrInv(predict(mylm, newdata=data.frame(X=pointX$date)))
# plot(yPred)
```

#### Another year: 2022

Probabilities point cloud

```{r}
year = '2022'
point22 <- chosenPoint %>% 
  filter(between(
    date, 
    as.Date(paste0(year, '-01-01')), 
    as.Date(paste0(year, '-12-31')))) %>% 
  drop_na(any_of(c("wet", "nat", "nonnat")))

point22X <- point22 %>% select(date) %>% 
  st_drop_geometry()
point22Y <- point22 %>% select(nat, nonnat, wet) %>% 
  st_drop_geometry() %>% 
  acomp()
plot(point22Y)
```

Simple linear fit Y ~ X

```{r}
simpleLm22 <- lm(ilr(point22Y) ~ X, data = data.frame(X = point22X$date))
summary(simpleLm22)
```

Model predictions
```{r}
plot(ilrInv(predict(simpleLm22)), pch=20)
```

Prediction residuals
```{r}
plot(ilrInv(resid(simpleLm22)))

# ellipses(mean(pointY), var(pointY))
# yPred <- ilrInv(predict(mylm, newdata=data.frame(X=pointX$date)))
# plot(yPred)
```

### Harmonic fit
Borrowing from Nick Clinton's YT tutorial on [Time Series Analysis in Earth Engine](https://www.youtube.com/watch?v=xQGNAYL7z80), construct a harmonic model.

#### Year 2022
Harmonic fit Y ~ X + cos(2piX) + sin(2piX)

See pg. 13 of [UsingCompositions.pdf](http://www.stat.boogaart.de/compositions/UsingCompositions.pdf) for a demonstration on how to build a linear model where the response variable is compositional data, and predictor variable/s is/are real data (which is the case here).

```{r}
xc <- cos(2*pi*point22$timeMillis) # cosine
xs <- sin(2*pi*point22$timeMillis) # sine
harmLm22 <- lm(
  ilr(point22Y) ~ X + XC + XS, 
  data = data.frame(X = point22X$date, XC = xc, XS = xs))
summary(harmLm22)
```

Model predictions
```{r}
plot(ilrInv(predict(harmLm22)), pch=20)
```

Prediction residuals
```{r}
plot(ilrInv(resid(harmLm22)))
```

